---
title: "NFL Betting 2024"
output: html_notebook
---

The goal of this project is to use historical NFL data to be able to predict future outcomes for sports betting.
It is currently a work in progress and is incomplete. 
So far, only initial data processing has been completed, and no models have been created yet.

Data from:
https://www.kaggle.com/datasets/tobycrabtree/nfl-scores-and-betting-data/

The original source files included some messy starter code.
I have used that starter code as inspiration, but have cleaned it up and made it more efficient by using dplyr.

Additional data may be needed to achieve meaningful results.
The original dataset may be lacking predictive power.

# Initialization
```{r}
library(dplyr)
library(stringr)
library(lubridate)

setwd("C:/Users/jknow/OneDrive/Documents/Sports Betting/NFL/2024")
```

# Load Data
```{r}
nfl <- read.csv("1 Data/spreadspoke_scores.csv", stringsAsFactors=F)
teams <- read.csv("1 Data/nfl_teams.csv",stringsAsFactors= F)
stadiums <- read.csv("1 Data/nfl_stadiums.csv",stringsAsFactors=F)
```

# Format Raw Data
```{r}
nfl$schedule_date <- as.Date(nfl$schedule_date, "%m/%d/%Y")
nfl$team_favorite_id <- as.factor(nfl$team_favorite_id)
```

# Add Game Info Columns
```{r}
# Add Home Team Columns
nfl <- nfl %>%
  left_join(teams,by=c('team_home'='team_name')) %>%
  mutate(team_home_name_short = as.factor(team_name_short),
         team_home_id = as.factor(team_id),
         team_home_id_pfr = as.factor(team_id_pfr),
         team_home_conference_current = as.factor(team_conference), # current conference for all rows
         team_home_division_current = as.factor(team_division), # current division for all rows
         team_home_conference_pre2002 = as.factor(team_conference_pre2002), # pre2002 conference for all rows
         team_home_division_pre2002 = as.factor(team_division_pre2002), # pre2002 division for all rows
         team_home_division = as.factor(ifelse(schedule_season < 2002,
                                               team_home_division_pre2002,
                                               team_home_division_current)),
         team_home_conference = as.factor(ifelse(schedule_season < 2002,
                                                 team_home_conference_pre2002,
                                                 team_home_conference_current))) %>%
  select(-team_name_short,-team_id,-team_id_pfr,-team_conference,-team_division,-team_conference_pre2002,-team_division_pre2002)

# Add Away Team Columns
nfl <- nfl %>%
  left_join(teams,by=c('team_away'='team_name')) %>%
  mutate(team_away_name_short = as.factor(team_name_short),
         team_away_id = as.factor(team_id),
         team_away_id_pfr = as.factor(team_id_pfr),
         team_away_conference_current = as.factor(team_conference), # current conference for all rows
         team_away_division_current = as.factor(team_division), # current division for all rows
         team_away_conference_pre2002 = as.factor(team_conference_pre2002), # pre2002 conference for all rows
         team_away_division_pre2002 = as.factor(team_division_pre2002), # pre2002 division for all rows
         team_away_division = as.factor(ifelse(schedule_season < 2002,
                                               team_away_division_pre2002,
                                               team_away_division_current)),
         team_away_conference = as.factor(ifelse(schedule_season < 2002,
                                                 team_away_conference_pre2002,
                                                 team_away_conference_current))) %>%
  select(-team_name_short,-team_id,-team_id_pfr,-team_conference,-team_division,-team_conference_pre2002,-team_division_pre2002)

# Get Game Unique ID
nfl <- nfl %>%
  mutate(game_id = gsub("-","",paste(as.Date(schedule_date,format='%m/%d/%Y'),team_away_id,team_home_id,sep="_")))

# Get Stadium Info
nfl <- nfl %>%
  left_join(stadiums,by=c('stadium'='stadium_name')) %>%
  mutate(stadium_type = as.factor(stadium_type))

# Get Schedule Info
season_length <- nfl %>%
  filter(schedule_playoff==FALSE) %>%
  group_by(schedule_season) %>%
  summarise(season_length = max(as.integer(schedule_week)))

# Create dummy variables for first week of season and last week of season [consider week after bye week, or short week ie, sunday>thursday game]
nfl <- nfl %>%
  left_join(season_length,by='schedule_season') %>%
  mutate(schedule_week_first = (schedule_week==1),
         schedule_week_last = (schedule_week==season_length))

# Add Day of Week Info
nfl <- nfl %>%
  mutate(schedule_day = wday(schedule_date, label=TRUE),
         schedule_month = month(nfl$schedule_date, label=TRUE),
         schedule_sunday = schedule_day%in%c("Sun"))
 
# Add Matchup Info
nfl <- nfl %>%
  mutate(conference_matchup = (team_home_conference==team_away_conference),
         division_matchup = (team_home_division==team_away_division))
```

# Add Betting Info
```{r}
# Add Spread and Over/Under Columns
nfl <- nfl %>%
  mutate(team_home_favorite = (as.character(team_favorite_id)==as.character(team_home_id)), # Doesn't exactly work because some rows have team_favorite_id %in% c("","PICK","LV"), which are not in team_home_id or team_away_id. Need to improve all lines that depend on team_home_favorite
         team_home_spread = ifelse(team_home_favorite,
                                   spread_favorite,
                                   -nfl$spread_favorite),
         team_away_spread = -team_home_spread,
         spread_type = as.factor(ifelse(team_home_spread==0,
                                        "Pick",
                                        ifelse(team_home_spread>0,
                                               "Home Underdog",
                                               "Home Favorite"))),
         spread_outlier = as.factor(ifelse(abs(spread_favorite)>14,
                                           "2TD+",
                                           ifelse(abs(spread_favorite)>10,
                                                  "1TD1FG+",
                                                  ifelse(abs(spread_favorite)>7,
                                                         "1TD+",
                                                         "No Outlier")))),
         over_under_outlier = as.factor(ifelse(over_under_line<33,
                                               "Under 2sd",
                                               ifelse(over_under_line<37,
                                                      "Under 1sd",
                                                      ifelse(over_under_line>50,
                                                             "Over 2sd",
                                                             ifelse(over_under_line>46,
                                                                    "Overer 1sd",
                                                                    "No Outlier"))))))

# Add More Game Columns
nfl <- nfl %>%
  mutate(tie = (score_home==score_away),
         team_winner = as.factor(ifelse(score_home>score_away,
                                        as.character(team_home_id),
                                        ifelse(score_home<score_away,
                                               as.character(team_away_id),
                                               "Tie"))),
         team_loser = as.factor(ifelse(score_home<score_away,
                                       as.character(team_home_id),
                                       ifelse(score_home>score_away,
                                              as.character(team_away_id),
                                              "Tie"))),
         team_home_result = as.factor(ifelse(score_home>score_away,
                                             "Win",
                                             ifelse(score_home<score_away,
                                                    "Loss",
                                                    "Tie"))),
         team_away_result = as.factor(ifelse(score_home<score_away,
                                             "Win",
                                             ifelse(score_home>score_away,
                                                    "Loss",
                                                    "Tie"))),
         score_total = score_home+score_away,
         team_home_win_count = ifelse(team_home_result=="Win",
                                      1,
                                      ifelse(team_home_result=="Loss",
                                             0,
                                             0.5)),
         team_away_win_count = ifelse(team_away_result=="Win",
                                      1,
                                      ifelse(team_away_result=="Loss",
                                             0,
                                             0.5)))

# Add Betting Result Columns
nfl <- nfl %>%
  mutate(over_under_result = as.factor(ifelse(score_total>over_under_line,
                                              "Over",
                                              ifelse(score_total<over_under_line,
                                                     "Under",
                                                     "Push"))),
         over_under_result_count = ifelse(over_under_result=="Over",
                                          1,
                                          ifelse(over_under_result=="Under",
                                                 0,
                                                 0.5)),
         spread_home_result = score_away-score_home,
         spread_away_result = -spread_home_result,
         score_favorite = ifelse(as.character(team_favorite_id)==as.character(team_home_id),
                                 score_home,
                                 ifelse(as.character(team_favorite_id)==as.character(team_away_id),
                                        score_away,
                                        0)),
         score_underdog = ifelse(as.character(team_favorite_id)==as.character(team_home_id),
                                 score_away,
                                 ifelse(as.character(team_favorite_id)==as.character(team_away_id),
                                        score_home,
                                        0)),
         spread_favorite_result = ifelse(as.character(team_favorite_id)==as.character(team_home_id),
                                         spread_home_result,
                                         ifelse(as.character(team_favorite_id)==as.character(team_away_id),
                                                spread_away_result,
                                                0)),
         spread_favorite_cover_result = as.factor(ifelse((score_favorite+spread_favorite)>score_underdog,
                                                         "Cover",
                                                         ifelse((score_favorite+spread_favorite)<score_underdog,
                                                                "Did Not Cover",
                                                                "Push"))),
         spread_favorite_cover_count = ifelse(as.character(spread_favorite_cover_result)=="Cover",
                                              1,
                                              ifelse(as.character(spread_favorite_cover_result)=="Did Not Cover",
                                                     0,
                                                     0.5)),
         spread_underdog_result = -spread_favorite_result,
         spread_underdog_cover_result = as.factor(ifelse((score_favorite+spread_favorite)>score_underdog,
                                                         "Did Not Cover",
                                                         ifelse((score_favorite+spread_favorite)<score_underdog,
                                                                "Cover",
                                                                "Push"))),
         spread_underdog_cover_count = ifelse(as.character(spread_underdog_cover_result)=="Cover",
                                              1,
                                              ifelse(as.character(spread_underdog_cover_result)=="Did Not Cover",
                                                     0,
                                                     0.5)),
         spread_home_cover_result = as.factor(ifelse(team_home_favorite,
                                                     as.character(spread_favorite_cover_result),
                                                     as.character(spread_underdog_cover_result))),
         spread_away_cover_result = as.factor(ifelse(!team_home_favorite,
                                                     as.character(spread_favorite_cover_result),
                                                     as.character(spread_underdog_cover_result))),
         spread_home_cover_count = ifelse(spread_home_cover_result=="Cover",
                                          1,
                                          ifelse(spread_home_cover_result=="Did Not Cover",
                                                 0,
                                                 0.5)),
         spread_away_cover_count = ifelse(spread_away_cover_result=="Cover",
                                          1,
                                          ifelse(spread_away_cover_result=="Did Not Cover",
                                                 0,
                                                 0.5)))
```

```{r}
# subset into games played versus future games
nflForecast <- subset(nfl,nfl$schedule_date > as.Date('2018-12-31')) # data for forecast games
nfl <- subset(nfl,nfl$schedule_date < as.Date('2018-12-31')) # data for played games
```

